Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-03-02 src/tex/ms.tex
Add corner plot for GW191105
new text for spin degeneracies

\subsection{Relation to inclination}
\label{sec:inclination}

@@

Since birefringence impacts the observed amplitude ratio of left- and right-handed modes, it could also affect inferences about the source inclination \cite{Alexander:2009tp}.
However, the two effects are degenerate only if the frequency dependence of Eq.~\eqref{eq:waveform_modification} is neglected.
This is easy to see from Eq.~\eqref{eq:waveform_modification}, since the implied polarization ratio for the $\ell = |m| = 2$ mode of a nonprecessing source is
\begin{equation}
    \frac{h_{L}^\mathrm{br}}{h_{R}^\mathrm{br}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2
    \exp\left({2\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\, \mathrm{Hz}}}\right)\, .
    \label{eq:modified_amplitude_ratio}
\end{equation}
For an isolated Fourier mode of definite frequency $f$, the effect of birefringence will be degenerate with a change in inclination; however, if multiple modes come into play, then no change in inclination can mask the effect of birefringence, which will affect the time domain waveform nontrivially (Fig.~\ref{fig:birefringence}).

@@

@@

\subsection{Result on GWTC-3}

% Violin plot
Figure \ref{fig:violin_kappa} shows the main result of our GWTC-3 analysis
as represented by the posterior distribution on $\kappa$ (abscissa) obtained for each event (ordinate). Posteriors are colored by the distance of the respective posterior mean from the origin in units of standard deviation, i.e., $|\mu_i/\sigma_i|$ for each event $i$.%
\footnote{We use the $i$ notation for individual events here to disambiguate from the population mean $\mu$ and standard deviation $\sigma$ of Eq.~\eqref{eq:posterior_of_mu_sigma}.}
The collective measurement obtained by assuming a shared $\kappa$ across events, Eq.~\eqref{eq:restricted_posterior}, is represented by its 90\%-credible symmetric interval (blue band) around the median (blue line);
this joint measurement is fully consistent with $\kappa = 0$ (dashed line), meaning we find no evidence for birefringence.

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}.

The two events that show the largest tension with $\kappa = 0$ is GW170818, for which $|\mu_i / \sigma_i| =$ \variable{output/GW170818_constraint.txt}, and GW200129\_065458 (henceforth GW200129), for which $|\mu_i / \sigma_i| =$ \variable{output/GW200129_constraint.txt}.
However, as we discuss further below, we have reason to think that the GW200129 preference for $\kappa < 0$ might be driven by noise anomalies in the Virgo detector; for that reason we consider the effect of excluding it from the joint result below.

\begin{table}[h]
    \caption{Events that best constrain $\kappa$, sorted by posterior standard deviation $\sigma_i$. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

Finally, a set of events stands out in Fig.~\ref{fig:violin_kappa} due to evident bimodality in the $\kappa$ posterior.
That is the case for the events listed in Table~\ref{tab:bimodal_events_mass}.
We understand this effect to be a consequence of the narrow sets of frequencies that are visible for these heavy events, as we mentioned in Sec.~\ref{sec:inclination} and further demonstrate below.

\begin{table}[h]
    \caption{Events that show bimodality in the $\kappa$ posterior tend to have high total masses. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

Figure \ref{fig:violin_kappa} shows certain variance in the distribution of $\kappa$ posteriors for different events, including some apparent outliers like GW170818.
This is not unexpected: assuming independent Gaussian noise instantiations for each event, we might expect up to ${\sim}3$ out of the 71 posteriors (i.e., ${\sim}5\%$) to deviate away from the true $\kappa$ value by over ${\sim}2\sigma$.

@@

% Case studies
\subsection{Notable events}

@@

% Case: GW170818
\subsubsection{GW170818}
GW170818 produced the posterior in most tension with $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows that this happens because birefringence opens up a region of parameter space with $\kappa>0$ for larger distances and smaller inclination angles than would be allowed in the GR case.
We can make sense of this because an edge-on, non-precessing source will produce linearly-polarized waves, so a smaller inclination will lead the two circular polarizations to have similar amplitudes.
On the other hand, having $\kappa > 0$ will enhance the left-handed modes during propagation.
The two effects can be balanced to yield the correct polarization ratio at the detector, as long as the distance is also enhanced to yield the right amount of birefringence and overall signal power.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
% Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
% This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
% We can see this effect in other events as well, but not as strongly as in GW170818.

@@

% Case: GW190521 (massive BBH)
\subsubsection{GW190521}
Even with the frequency-dependent birefringence model, some events still have a bimodal $\kappa$ distribution.
Consider GW190521, the most massive binary black hole merger in the events we included.
The degeneracy between $\kappa$ and $\iota$ cannot be broken by the frequency-dependent birefringence model, as shown in Fig.~\ref{fig:corner_GW190521}.
The reason could be this GW event's frequency range is too narrow to break the degeneracy.
The effect of birefringence at different frequencies within the range is similar, so the degeneracy cannot be broken.
Therefore, the narrow frequency range can be a possible reason why the $\kappa$ distribution of GW190521 is bimodal.
Note that $\kappa=0$ is still within the $90\%$ confidence interval, with a non-negligible probability.
The bimodal $\kappa$ distribution only means that the degeneracy between $\kappa$ and $\iota$ cannot be broken in this case.  \wf{You should still explain \emph{why} the model seems to prefer $\left| \kappa \right| > 0$.  I \emph{think} the reason here is the distance prior---there is a lot more ``phase space'' at large distances that can be accessed by allowing $\kappa$ to be non-zero and choosing a corresponding inclination so that the amplification effects of birefringence are compensated for by the larger distance compared to the pure-GR run.  Does this make sense to you?  If so, can you write it up in this paragraph?}

@@

% Case: GW150914 (example of broken degeneracy)
\subsubsection{Result on GW150914}
In this study, we included frequency dependence of birefringence, which would affect the posterior of $\kappa$ obtained from \ac{PE}.
Consider GW150914, the first GW detected by LIGO, as an example.
In Fig.~\ref{fig:corner_GW150914}, we show the posteriors of the parameters of GW150914.
With the frequency-independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
% explain the kappa gap

@@

>>>>

\subsection{Inclination and other degeneracies}
\label{sec:inclination}

@@

Since birefringence impacts the observed amplitude ratio of left- and right-handed modes, it could also affect inferences about the source inclination \cite{Alexander:2009tp}.
However, the two effects are degenerate only if the frequency dependence of Eq.~\eqref{eq:waveform_modification} is neglected.
This is easy to see from Eq.~\eqref{eq:waveform_modification}, since the implied polarization ratio for the $\ell = |m| = 2$ mode of a nonprecessing source is
\begin{equation}
    \frac{h_{L}^\mathrm{br}}{h_{R}^\mathrm{br}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2
    \exp\left({2\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\, \mathrm{Hz}}}\right)\, .
    \label{eq:modified_amplitude_ratio}
\end{equation}
For an isolated Fourier mode of definite frequency $f$, the effect of birefringence will be degenerate with a change in inclination; however, if multiple modes come into play, then no change in inclination alone can mask the effect of birefringence, which will affect the time domain waveform nontrivially (Fig.~\ref{fig:birefringence}).

@@

Besides source inclination, the effect of birefringence can be (partially) degenerate with other parameters---most notably, precession, as we find in Sec.~\ref{sec:Results}.
Indeed, under the right conditions, the frequency-dependent amplification/dampening of polarizations caused by $\kappa$ can be effectively absorbed by changes in intrinsic parameters, like the spins;
in that case, the degeneracy with inclination may appear even for lighter \acp{BBH} (although usually to a lesser degree than heavy binaries).
We chart some of these degeneracies as part of the results we present in Sec.~\ref{sec:Results}.
% In particular, we find that, under certain conditions, the frequency-dependent birefringence of Eq.~\eqref{eq:waveform_modification} can mimic the time-dependent amplitude modulation expected from a precessing system.

@@

\subsection{GWTC-3 result}

% Violin plot
Figure \ref{fig:violin_kappa} shows the main result of our GWTC-3 analysis
as represented by the posterior distribution on $\kappa$ (abscissa) obtained for each event (ordinate). Posteriors are colored by the distance to the origin of the respective mean in units of standard deviation, i.e., $|\mu_i/\sigma_i|$ for each event $i$.%
\footnote{We use the $i$ notation for individual events here to disambiguate from the population mean $\mu$ and standard deviation $\sigma$ of Eq.~\eqref{eq:posterior_of_mu_sigma}.}
The collective measurement obtained by assuming a shared $\kappa$ across events, Eq.~\eqref{eq:restricted_posterior}, is represented by its 90\%-credible symmetric interval (blue band) around the median (blue line);
this joint measurement is fully consistent with $\kappa = 0$ (dashed line), meaning we find no evidence for birefringence.

Figure \ref{fig:violin_kappa} makes it clear that not all \acp{BBH} in GWTC-3 are equally informative about birefringence.
When considered individually, the events that best constrain $\kappa$ are listed in Table~\ref{tab:best_events_kappa}, in order of increasing standard deviation $\sigma_i$.
That table also shows the credible level (CL) at which the posterior supports $\kappa = 0$, whereby $\mathrm{CL} = 0$ ($\mathrm{CL} = 1$) means the posterior supports that value with high (low) probability.

Judging by $\mu_i/\sigma_i$, the two events that show the largest tension with $\kappa = 0$ are GW170818, for which $\mu_i / \sigma_i =$ \variable{output/GW170818_constraint.txt}, and GW200129\_065458 (henceforth GW200129), for which $\mu_i / \sigma_i =$ \variable{output/GW200129_constraint.txt}.
However, as we discuss in Sec.~\ref{sec:GW200129}, we have reason to think that the preference for $\kappa < 0$ in GW200129 might be driven by noise anomalies in the Virgo detector; for that reason, in the next section we consider the effect of excluding this event from the joint result (we find its impact to be minimal).

\begin{table}
    \caption{Events that best constrain $\kappa$, sorted by posterior standard deviation $\sigma_i$. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

Finally, a set of events stands out in Fig.~\ref{fig:violin_kappa} due to evident bimodality in the $\kappa$ posterior.
To a varying degree, that is the case for those events listed in Table~\ref{tab:bimodal_events_mass}, which tend to have quite high total masses (Table~\ref{tab:bimodal_events_mass} show total mass as measured in the standard  \ac{GR} analysis).
\mi{THOMAS: is this correct? do these numbers come from the GR runs? they should}
For these bimodal posteriors, $\mu_i/\sigma_i$ is not a good proxy for agreement with \ac{GR}; instead, we can rely on $\mathrm{CL}(\kappa = 0)$.
By this measure, the bimodal events are some of the least consistent with $\kappa = 0$, GW190521 in particular. (However, note that configurations with $\kappa = 0$ for this event are still within the higher dimensional 90\%-credible region, when other parameters are considered and not just the one-dimensional marginal; we discuss this in Sec.~\ref{sec:GW190521}.)

As we mentioned in Sec.~\ref{sec:inclination}, we understand the bimodality in $\kappa$ to be primarily a consequence of the narrow set of frequencies that are visible for these heavy events, which makes the effect of birefringence (partially) degenerate with a change in inclination.
Other parameter degeneracies also come into play, especially for the lighter events GW191105\_143521 and GW170104.
We discuss this in detail in a dedicated section below (Sec.~\ref{sec:notable}).

\begin{table}
    \caption{Events that show bimodality in the $\kappa$ posterior tend to have high total masses. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

Figure \ref{fig:violin_kappa} shows certain degree of variance in the distribution of $\kappa$ posteriors for different events, including some apparent outliers like GW170818.
This is not unexpected: assuming independent Gaussian noise instantiations for each event, we might expect up to ${\sim}3$ out of the 71 posteriors (i.e., ${\sim}5\%$) to deviate away from the true $\kappa$ value by over ${\sim}2\sigma$.

@@

% Case studies
\subsection{Notable events}
\label{sec:notable}

@@

% Case: GW170818
\subsubsection{GW170818}
\label{sec:GW170818}

GW170818 produced the posterior in most tension with $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows that this happens because birefringence opens up a region of parameter space with $\kappa>0$ for larger distances and smaller inclination angles than would be allowed in the GR case.
We can make sense of this because an edge-on, non-precessing source will produce linearly-polarized waves, so a smaller inclination will lead the two circular polarizations to have similar amplitudes.
On the other hand, having $\kappa > 0$ will enhance the left-handed modes during propagation.
The two effects can be balanced to yield the correct polarization ratio at the detector, as long as the distance is also enhanced to yield the right amount of birefringence and overall signal power.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
% Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
% This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
% We can see this effect in other events as well, but not as strongly as in GW170818.

@@

% Case: GW190521 (massive BBH)
\subsubsection{GW190521}
\label{sec:GW190521}

Even with the frequency-dependent birefringence model, some events still have a bimodal $\kappa$ distribution.
Consider GW190521, the most massive binary black hole merger in the events we included.
The degeneracy between $\kappa$ and $\iota$ cannot be broken by the frequency-dependent birefringence model, as shown in Fig.~\ref{fig:corner_GW190521}.
The reason could be this GW event's frequency range is too narrow to break the degeneracy.
The effect of birefringence at different frequencies within the range is similar, so the degeneracy cannot be broken.
Therefore, the narrow frequency range can be a possible reason why the $\kappa$ distribution of GW190521 is bimodal.
Note that $\kappa=0$ is still within the $90\%$ confidence interval, with a non-negligible probability.
The bimodal $\kappa$ distribution only means that the degeneracy between $\kappa$ and $\iota$ cannot be broken in this case.  \wf{You should still explain \emph{why} the model seems to prefer $\left| \kappa \right| > 0$.  I \emph{think} the reason here is the distance prior---there is a lot more ``phase space'' at large distances that can be accessed by allowing $\kappa$ to be non-zero and choosing a corresponding inclination so that the amplification effects of birefringence are compensated for by the larger distance compared to the pure-GR run.  Does this make sense to you?  If so, can you write it up in this paragraph?}

@@

% Case: GW150914 (example of broken degeneracy)
\subsubsection{GW150914}
In this study, we included frequency dependence of birefringence, which would affect the posterior of $\kappa$ obtained from \ac{PE}.
Consider GW150914, the first GW detected by LIGO, as an example.
In Fig.~\ref{fig:corner_GW150914}, we show the posteriors of the parameters of GW150914.
With the frequency-independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
% explain the kappa gap

@@

\section{placeholder for GW191105}

\begin{figure}[h]
    \script{corner_GW191105.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW191105.pdf}
    \caption{
        Corner plot of GW191105.
    }
    \label{fig:corner_GW191105}
\end{figure}
